---
title: "Assignment 1"
author: "gg"
date: "September 17, 2015"
output: html_document
---
# Assignment 1
Please hand this in as an .Rmd document, and I will convert it to html or pdf for marking

You will be graded for your comments, I need to know that you understand the basics of R and how the code is working. You are really writing for your future self, so it should be easy to separate the code from the comments. Place general comments about what you are trying to accomplish outside of the R code itself.

I know that part 4 will be a challenge, but it can be done simply using a for loop. You first have to decide exactly what it is you want to do---this is almost always the most difficult part.

## functions and loops in R

loops and functions are standard fare in R programs. Another is the if-else statement that is used to do something conditionally. For example:

explanation of if-else function in R is here:
http://www.programiz.com/r-programming/if-else-statement

#### part 1: commment this code. Make sure to use your own wording
```{r}
for(i in 1:10){
	if ( i %% 2 == 0) {
		print( c(i, "is even"))
	} else if (i %% 2 != 0) {
		print(c(i, "is odd"))
	}
}
```

#### part 2: change the code to print the value of i and 'huzzah' if the value of i is a prime number for numbers between 10 and 20. Again comment the code in your own words.

## P values

#### part 3: what is the difference in interpretation between a P value, a family wide error-rate value determined by the Bonferroni correction and a false discovery rate determined by the Benjamini-Hochberg method?

#### part 4: explore part 4 below, what can you conclude about the ability of an experiment to see read differences it the dataset has 10000 genes, but only 6-10 samples equally divided between control and experimental conditions?

```{r}
# I have changed the variable assignmet to x.mat from class so that you can do some exploratory analysis.

# variables for the means in group 1 and 2
mean1 <- 10
mean2 <- 10
sd <- 2 # we will hold this constant
sample.size <- 6 # make sure this is always even!!

#### should not need to change from here
x.mat <- matrix(data=NA, nrow=100, ncol=sample.size)

max.set1 <- ncol(x.mat)/2
min.set2 <- ncol(x.mat)/2+1
# was very annoying to get the bracketing just right, so don't mess with this
for(i in 1:nrow(x.mat)){
  x.mat[i,1:1:max.set1] <- rnorm((ncol(x.mat)/2), mean1, sd)
  x.mat[i,min.set2:ncol(x.mat)] <- rnorm((ncol(x.mat)/2), mean2, sd)
}
#### to here

# provided for reference as to how to grab the P values and get the mean
tt <- apply(x.mat, 1, function(x){as.numeric(t.test(x[1:max.set1], x[min.set2:ncol(x.mat)])[3])})
mean(tt)
```

Using this dataset, and the code above as a starting point, change the mean and sample sizes to determine the effect of sample size and difference between means on P values. Present your data as a graph of the mean P value obtained for 100 replicates (nrow in x.mat). Try varying the difference in means from 0-4 fold, and the sample size from 6-100. Do both in increments (i.e., you do not need to do every value since we are looking for trends)

It is acceptable to do this manually, and to provide a table, however full marks will be given if the analysis can be done inside as a loop.


