---
title: "final assignment"
author: "gg"
date: "`r Sys.Date()`"
fig_caption: true
output: pdf_document
---

To run this file:
Rscript -e "rmarkdown::render('final.Rmd')"

Exploratory data analysis to determine if  replication is tight in the Bottomly dataset

try a couple of scatter plots, and then a biplot

```{r setup, echo=FALSE, message=FALSE}
# load the compositional framework
library(compositions)
library(zCompositions)
library(ALDEx2)
source("~/git/compositions/CoDaMicro/chunk/codaMB_functions.R")

# this is the metadata
meta <- read.table("~/Documents/LaTeX/445/9545_fall2015/final_R/bottomly_phenodata.txt", row.names=1,
    header=T)

# this is the table of counts
d <- read.table("~/Documents/LaTeX/445/9545_fall2015/final_R/bottomly_count_table.txt", header=T, row.names=1)

```
first reduce the dataset since there are `r nrow(d)` genes. This is a lot, and there is no need to keep them all.

```{r biplot, echo=TRUE, message=FALSE, warning=FALSE}
# reduce the dataset
# keep genes that are in at least 40% of the samples
# keep genes that with a frequency across any sample of at least 1/1e-6
 d.filter <- codamb.filter(d, min.prop=1e-6, min.fraction=0.4, samples.by.row=FALSE)

# clr transform the data with a uniform prior(rough first pass)
d.clr <- t(apply(d.filter+0.5,2, function(x){log(x) - mean(log(x))}))
pcx <- prcomp(d.clr)
mv <- mvar(d.clr)
# basically only two components
plot(pcx$sdev^2/mv, type="b")

# the samples are very clumpy and
# the samples can be seen to cluster by experiment number
# experiment 4 is very different than experiment 6 and 7
#
plot(pcx$x[,1], pcx$x[,2], pch=paste(meta$experiment.number, sep=""), xlab=round(pcx$sdev[1]^2/mv,2), ylab=round(pcx$sdev[2]^2/mv,2))

# we can also plot the contribution of each sample to the mvar and we see that both the groups are bimodal

# so yes, there is a lot of structure in the data that is not anticipated and will confound later analyses. The best thing to do would be to only use the 6/7 sample sets. They are much tighter. We will persevere with the analysis as is.
```
The reduced dataset contains `r nrow(d.filter)`, which is more manageable and really should not decrease power that much.

On to phi!

```{r phi, echo=TRUE}

```
