---
title: "9545 2018 Day 2"
output: html_document
---
# Review of day 1



# Functions in R

Functions in R allow us to write a method once and use it over and over again. All commands in R are functions. Functions can operate on data or on other functions. Understanding how to make a function will help you understand how to call a function and use the built-in functions effiently.

Functions have the form `function.name <- function(param){do.something(param)}`. As an example, let's make a function to do the plotting for the Anscombe datastet.

```{r, echo=T}
# define the function
plot.anscombe.gg <- function(dataset, off.set, col.fill){

    nm.txt <- paste("Quartet ",off.set, sep="")
    cor.txt <- paste("corr =", cor.list[[off.set]], sep="")
    plot(dataset[,off.set], dataset[,off.set+4] , main=nm.txt, sub=cor.txt,
        xlab=paste("x",off.set), ylab=paste("y",off.set), pch=21, col="darkgrey", bg=col.fill[i])
    abline(cor.list[[off.set+4]])

}
```

And this simplifies our for loop a lot!

Add comments to the two code blocks

```{r, echo = T}
col.fill=c("red", "yellow", "green","blue")

par(mfrow=c(2,2))

for(i in 1:4){ plot.anscombe.gg(anscombe, i, col.fill) }

```

## Here is another function that generates random data along a line

```{r, echo=T}
# function to make random data on a line
cor.scatter <- function(off.set){
    x.r <- runif(100,off.set,off.set + 10)
    y.r <- vector()
    for( i in 1:length(x.r)){
        y.r[i] <- rnorm(1, mean=x.r[i], sd=4)
    }
    return(cbind(x.r,y.r))
}

# the data are saved as 2 column matrices
# random data
s1 <- cor.scatter(10)
s2 <- cor.scatter(1)
s3 <- cor.scatter(20)
s.all <- rbind(s1,s2,s3)

# try
# make this a list

# plot it
par(mfrow=c(1,1))
plot(s.all[,1], s.all[,2])
points(s1[,1], s1[,2], col="red")
points(s2[,1], s2[,2], col="blue")
points(s3[,1], s3[,2])

# what is the correlation in each?
# add a line of best fit for each and for the whole

# now translate and retry
s2.t <- cbind(s2[,1] + 20, s2[,2])
s3.t <- cbind(s3[,1] - 20, s3[,2])
s.all.t <- rbind(s1, s2.t, s3.t)
par(mfrow=c(1,2))
plot(s.all.t[,1], s.all.t[,2])
points(s1[,1], s1[,2], col="red")
points(s2.t[,1], s2.t[,2], col="blue")
points(s3.t[,1], s3.t[,2])
```

# Writing and reading data

Here we will learn two ways to save data. One in a plain text file, the other as an R data structure more plotting.

The simplest way is to save and load the data. This is perfect for multidimensional data and for lists.  For example:

```{r, echo=T}
# what is happening here?
save(s.all.t, file="s.all.t")
rm(s.all.t)
load("s.all.t")
```

The second way is to save the data as a table. This is perfect for passing around to your friends, or for supplementary info in a paper.

```{r, echo=T}

write.table(s.all.t, file="s.all.t.txt", sep="\t", quote=F, col.names=F)

jnk <- read.table("s.all.t.txt", sep="\t", header=F, row.names=1, stringsAsFactors=F)
