---
title: "9545 2018 Days 3-4"
output: html_document
---
# Exploratory Data Analysis
### AKA Look at your Data!!!!

We will use a dataset from the `ALDEx2 R` package. There are many datasets available in R. We have seen and used the `anscombe` dataset built into base R. Now we will explore a real dataset derived from high throughput sequencing (actually a MiSeq, so kinda low end of HTS).

```{r, echo=T}
# load the R package
library(ALDEx2)

# Load up the data
data(selex)
```

At this point we have a dataset that we can explore.

- what is the number of rows and columns of the data?

- what type of data is this?

- what are the row and column characteristics?
    -four ways to get the sums of the columns

```{r, echo=T}
# the old fashioned way, one column at a time
sum(selex[,1])

# using apply() to find sum of each column
apply(selex, 2, sum)

# a for loop (always the last resort)
for(i in ncol(selex)){sum(selex[,i])}

# the best way because is vectorized
colSums(selex)
```
looking at distributions

```{r, echo=T}
#linear space
# counting
par(mfrow=c(2,7))
apply(selex, 2, function(x) hist(x, breaks=99))

# log space
# relative (exponential values, fold changes)
par(mfrow=c(2,7))
apply(selex, 2, function(x) hist(log10(x + 0.5), breaks=99))

```

similarity between samples

```{r, echo=T}
# scatterplots are your friend
par(mfrow=c(1,1))
plot(selex[,1], selex[,7], pch=19, col=rgb(0,0,0,0.1))
abline(0,1, col="red")
cor(selex[,1], selex[,7]

par(mfrow=c(1,1))
plot(selex[,1]+0.5, selex[,7]+0.5, pch=19, col=rgb(0,0,0,0.1), log="xy")
abline(0,1, col="red")
```
